% ! TeX root = ../main.tex

% \section{Requirements for Closed-Loop Observability}
\section{Requirements}
\label{sec:orca:req}

Our goal is to accelerate the full spectrum of observability workflows, from offline trace analysis
to real-time diagnosis.
This requires analytics to transform fine-grained telemetry in situ, terminating in aggregated
views or detailed traces but reorganized for efficient offline analyses.
Additionally, control semantics are needed for interventions ranging from steering application views to steering the
application itself.
We model BSP as consisting of timesteps punctuated by synchronous collectives~\cite{:MPI,:NCCL},
with timesteps containing one or more collectives.
Steerable observability for this model must support three key requirements.


\pplushead{BSP-Aware Semantic Guarantees}
Both the data and control paths must be synchronization-aware.
On the data path, both transport and analytics must offer consistency guarantees, as all telemetry
for a timestep must be collected and analyzed as a single unit.
Control inputs must similarly be applied consistently and atomically at timestep boundaries
across all ranks.
While uncoordinated updates may be tolerable in some cases, they break correctness for
interventions such as hyperparameter changes --- a timestep-consistent control plane generalizes
across use cases.

\pplushead{Low-Cost, Low-Latency Feedback}
For always-on view-driven observability to be viable, overhead must be low and proportional to view
complexity.
Induced jitter must be minimized, so that continuously monitored aggregates incur minimal cost,
while detailed traces remain available on demand.
Analytics must execute in real-time, as feedback latency bounds intervention latency.


\pplushead{Declarative and Flexible Analytics}
No predefined set of analyses can anticipate every pathology.
Operators must be able to specify views and collection policies via a programmable interface at
runtime, without recompilation or restart.
Collection and analytics must be schema-agnostic, enabling visibility into any layer of the runtime
stack --- from MPI and fabric libraries to the application itself.
\vspace{0.2em}

% Supporting these requirements at scale requires balancing competing constraints ---
% distributed execution for scalability, asynchronous operation for low overhead, and coordinated semantics for correctness ---
% all the while keeping analytics expressive enough for ad-hoc queries.
% At scale, distributed execution becomes necessary, creating competing constraints between asynchrony for low overhead
% and coordination for correctness.
At scale, both data and control paths necessitate distributed execution, requiring a careful balance of consistency guarantees
with asynchrony for low overhead.
Our design leverages modern columnar primitives (Arrow~\cite{:ApacheArrow}, Parquet~\cite{:ApacheParquet}),
an extensible query engine called DataFusion~\cite{Lamb2024:DataFusion}, and RDMA for low-overhead transport,
with a tree-based topology for reductions and broadcast.
% At scale, these requirements pose competing constraints over distributed execution --- correctness necessitates
% coordination across a large analytics tier, while
% aligning analytics and control with a tree-based reductions and broadcasts respectively.
% DataFusion~\cite{Lamb2024:DataFusion}) with RDMA for low-overhead transport, aligning data
% partitioning with timestep boundaries and analytics with a reduction topology.

% work interspersed by global synchronization via collectives~\cite{:MPI,:NCCL}.
% On the data path, this implies coordinated collection and analytics.
% BSP observability requires guarantees on both data and control paths.
% On the data path, this means coordinated collection: detecting the straggler at a barrier requires
% telemetry from all ranks for that barrier instance.
% Sampling yields uncoordinated subsets insufficient for root cause attribution.
% Control path semantics must similarly align with workload structure.

