% 'nonacm' to remove any running header
\documentclass[nonacm,sigconf]{acmart}

\usepackage{cite}
\usepackage{amsmath,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[svgnames]{xcolor}
\usepackage{booktabs}
\usepackage{tagging}
\usepackage{enumitem}
\usepackage{minted}

\usepackage{sc25repro}
\pagenumbering{gobble} % no page numbers
\pagestyle{empty} % suppress acmart running header (no \title/\author set)
\makeatletter\def\shorttitle{}\def\@shorttitle{}\makeatother

\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

\newcommand{\profmini}{\textsf{Mini}}
\newcommand{\profdet}{\textsf{Detailed}}
\newcommand{\textmpiw}{\texttt{MPI\_Wait}}

\newmintinline[code]{text}{bgcolor=Gainsboro!40}

% --- environments for verbatim blocks ---
\newenvironment{bash}
{\VerbatimEnvironment\begin{minted}[frame=lines,framesep=2mm,baselinestretch=0.9,bgcolor=WhiteSmoke,fontsize=\footnotesize]{bash}}
{\end{minted}}

\newenvironment{example}
{\VerbatimEnvironment\begin{minted}[bgcolor=GhostWhite,fontsize=\small]{text}}
{\end{minted}}

% Uncomment/comment the following usetag lines
% if you would like to get an explanation or an example.
% Don't forget to comment them for the final submission.
\usetag{explanation}
\usetag{example}

\begin{document}

\twocolumn[%
	{\begin{center}
				\Huge
				Appendix: Artifact Description/Artifact Evaluation
			\end{center}}
]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  AD Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendixAD

\section{Overview of Contributions and Artifacts}

\subsection{Paper's Main Contributions}

\begin{description}
	\item[$C_1$] \textbf{Low-overhead tracing.} Always-on conventional tracing at \textasciitilde0.8--2\% overhead (vs 18--81\% for TAU/dftracer), with Parquet traces 3.6--44$\times$ smaller than other formats. Figs.~5, 6, 8.
	\item[$C_2$] \textbf{Faster post-mortem analysis.} ORCA traces enable post-mortem queries multiple orders of magnitude faster than other analytics-oriented tracers. Fig.~7.
	\item[$C_3$] \textbf{In-situ analytics.} Operator pushdown lets MPI-tier flows run at sub-2\% overhead while cutting data movement out of MPI ranks by 98.5--99.999\%. Fig.~9 and Table~1. \emph{Not supported by conventional tracers.}
	\item[$C_4$] \textbf{Agentic debugging.} Steerable telemetry lets an LLM agent localize a months-long 4096-rank AMR anomaly in 27 minutes, with 144$\times$ volume reduction when reverting to baseline. Fig.~10. \emph{Not supported by conventional tracers.}
\end{description}

\subsection{Computational Artifacts}

\begin{description}
	\item[$A_1$] ORCA-Augmented AMR Code
	\item[$A_2$] Post-Mortem Query Scripts
	\item[$A_3$] LLM Prompt for Agentic Workflow
\end{description}

\begin{center}
	\begin{tabular}{rll}
		\toprule
		Artifact ID & Contributions       & Related                                \\
		            & Supported           & Paper Elements                         \\
		\midrule
		$A_1$       & $C_1$, $C_3$, $C_4$ & Figs.~5, 6, 8 ($C_1$)                  \\
		            &                     & Fig.~9 and Table~1 ($C_3$)             \\
		            &                     & Fig.~10, live system for $A_3$ ($C_4$) \\
		\midrule
		$A_2$       & $C_2$               & Fig.~7                                 \\
		\midrule
		$A_3$       & $C_4$               & Fig.~10                                \\
		\bottomrule
	\end{tabular}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Artifact Identification}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% ----------------------------------------------
% A1 --- ORCA-Augmented AMR Code
% ----------------------------------------------
\newartifact

\artrel

$A_1$ is the instrumented BSP workload --- Phoebus + Parthenon's Sedov Blast Wave 3D AMR code linked against ORCA --- used as the common substrate for three contributions:

\begin{itemize}
	\item $C_1$ (Figs.~5, 6, 8): running $A_1$ under different tracer configurations produces the runtime overhead, trace size, and architectural ablation measurements.
	\item $C_3$ (Fig.~9): running $A_1$ with the in-situ flows from Table~1 produces the pushdown overhead and data-reduction measurements.
	\item $C_4$ (Fig.~10): $A_1$ is the live system that $A_3$'s LLM agent attaches to and steers.
\end{itemize}

The artifact is provided with two profiles: \profmini{} and \profdet{}.

\begin{itemize}
	\item \profmini{}: triggers short runs, sufficient to verify artifact function.
	\item \profdet{}: requires 262 nodes on the PDL Wolf cluster and triggers longer experiments, necessary to replicate all results.
\end{itemize}

\artexp

\begin{itemize}
	\item \profmini{}: three timesteps of a simple example program execute, generating Parquet traces via ORCA.
	\item \profdet{}: 2000 timesteps of the Blast Wave AMR code, generating Parquet traces via ORCA in two modes---lightweight and detailed. 
	A baseline run is also triggered for overhead comparison.
\end{itemize}

\arttime

\begin{itemize}
	\item Setup: \textasciitilde30 minutes (building the entire project)
	\item Execution: \textasciitilde1 min for \profmini{}, \textasciitilde2--3 hours for \profdet{} (Baseline and ORCA only)
	\item Analysis: $<$10 minutes
\end{itemize}

\artin

\artinpart{Hardware}

\begin{itemize}
	\item \profmini{}: one multi-core Linux node
	\item \profdet{}: 262 nodes on the PDL Wolf cluster
\end{itemize}

\artinpart{Software}

Required dependencies, must be present in the build environment.
Versions used in evaluation environment are noted in parentheses. 
Other environments may necessitate tweaks to build parameters etc.

\begin{itemize}
	\item A Linux environment (Ubuntu 22.04)
	\item An MPI implementation (MVAPICH2 2.3.7)
	\item C/C++ compiler (GCC 11.4)
	\item CMake (3.22)
	\item Rust toolchain (cargo/rustc, stable 1.92)
	\item TAU (2.34.1)
\end{itemize}

\textbf{Dependencies built by orca-umbrella} (\url{https://github.com/pdlfs/orca-umbrella})

\vspace{0.5em}
\noindent \emph{ORCA-Augmented AMR Code:}
\begin{itemize}[topsep=0pt]
	\item ORCA --- \url{https://github.com/pdlfs/orca}
	\item Phoebus --- \url{https://github.com/lanl/phoebus}
	\item Parthenon --- \url{https://github.com/parthenon-hpc-lab/parthenon}
	\item amr-tools --- \url{https://github.com/pdlfs/amr-tools}
	\item Kokkos 4.0.01 --- \url{https://github.com/kokkos/kokkos}
	\item HDF5 1.12.2 --- \url{https://github.com/HDFGroup/hdf5}
\end{itemize}

\vspace{0.5em}
\noindent\emph{ORCA runtime / data-path:}
\begin{itemize}[topsep=0pt]
	\item Apache Arrow 19.0.0 --- \url{https://github.com/apache/arrow}
	\item Snappy --- \url{https://github.com/google/snappy}
	\item Mercury --- \url{https://github.com/mercury-hpc/mercury}
	\item libfabric --- \url{https://github.com/ofiwg/libfabric}
	\item DuckDB v1.2.1 --- \url{https://github.com/duckdb/duckdb}
	\item yaml-cpp --- \url{https://github.com/jbeder/yaml-cpp}
	\item PSM (forked) --- \url{https://github.com/pdlfs/psm}
\end{itemize}

\vspace{0.5em}
\noindent\emph{Comparison tracers (for complete results replication):}
\begin{itemize}[topsep=0pt]
	\item DFTracer v2.0.2 (forked) --- \url{https://github.com/anku94/dftracer}
	\item Score-P 9.3 (patched, patch automatically applied during build) --- \url{https://www.vi-hps.org/projects/score-p/}
	\item Caliper (patched, patch automatically applied during build) --- \url{https://github.com/LLNL/Caliper}
\end{itemize}

\artinpart{Datasets / Inputs}

\begin{itemize}
	\item AMR decks: \code{blast_wave_3d.<nranks>.pin} (nranks in 512, 1024, 2048, 4096)
	\item \code{wfopts.yml}: tuned config for Wolf cluster, only needed for \profdet{} profile
\end{itemize}

\artinpart{Installation and Deployment}

Preparation commands (indicative, from Ubuntu 22.04, may vary across environments):

\begin{bash}
# Install basic utilities
sudo apt install -y gcc g++ make cmake autoconf \
automake libtool pkg-config git

# Install rust toolchain via rustup (non-interactive, stable)
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs \
| sh -s -- -y --default-toolchain stable
. "$HOME/.cargo/env"
\end{bash}

Now, orca-umbrella can be built, and will also build the rest of the dependencies in-tree.

\begin{bash}
# Install prefix should be a shared dir on the cluster
OR_PREFIX=/tmp/orca-prefix
OR_BUILD=$OR_PREFIX/build
OR_INSTALL=$OR_PREFIX/install
mkdir -p $OR_PREFIX $OR_BUILD $OR_INSTALL

git clone https://github.com/pdlfs/orca-umbrella.git $OR_PREFIX
cd $OR_BUILD
cmake -DCMAKE_INSTALL_PREFIX=$OR_INSTALL \
	$OR_PREFIX/orca-umbrella
make -j<nproc>
\end{bash}

\noindent Once successfully built, all artifacts will be available in \code{$OR_INSTALL}. 
Run the artifact using scripts in \code{$OR_INSTALL/scripts}.

\begin{itemize}
	\item \profmini{}: run \code{run_amr_test.sh}
	\item \profdet{}: Generate a valid hostfile as per MPI configuration, with at least 262 nodes, and run:
			\begin{bash}
OR_HOSTFILE=/path/to/hostfile \
	$OR_INSTALL/scripts/run_orca_detailed.sh
			\end{bash}
\end{itemize}

\artcomp

\begin{itemize}
	\item The relevant MPI codes are executed under exhaustive MPI/Kokkos trace collection modes, as per the configured scales and timesteps.
	\item \profdet{}: Executes a baseline run and ORCA in lightweight and detailed modes at different scales, and reports overhead.
\end{itemize}

\artout

Each run yields timestep-partitioned Parquet traces in the job directory, organized by schema:

\begin{example}
parquet/
|-- kokkos_events/
|   `-- ts=<timestep_range>/
|       `-- ranks=<rank_range>.parquet
|-- mpi_collectives/
|   `-- ...
|-- mpi_messages/
|   `-- ...
`-- orca_events/
\end{example}

\noindent \profdet{}: Wall-clock times are compared with baseline, and measured runtime overhead of different modes is reported.

% ----------------------------------------------
% A2 --- Post-Mortem Query Scripts
% ----------------------------------------------
\newartifact

\artrel

$A_2$ is the post-mortem analytics suite that executes a 
query suite consisting of three queries, against traces captured by running $A_1$.
The three queries in the suite are: outlier waits, outlier collectives, and timestamp range.
It supports $C_2$ by producing the query-latency comparison in Fig.~7.

\artexp

The query suite runs against all available traces. Row counts in executed queries are displayed for verification and time taken by queries is logged as a CSV.

\arttime

\begin{itemize}
	\item Setup: $<$5 minutes
	\item Execution: $<$1 minute (ORCA only)
	\item Analysis: $<$1 minute
\end{itemize}

\artin

\artinpart{Hardware}
One multi-core Linux node.

\artinpart{Software}
ORCA trace queries: Python 3 environment with Polars.
\begin{itemize}
	\item Python (3.12.11)
	\item Polars (1.39.3)
\end{itemize}

\artinpart{Datasets / Inputs}
ORCA traces generated by the AMR codes in $A_1$.

\artinpart{Installation and Deployment}
All Python scripts will be copied to \code{$INSTALL_TREE/scripts/orcaquery} during the orca-umbrella build (see A1). 
Polars may be installed by running
\begin{center}
\code{pip install -r requirements.txt}
\end{center}
(or equivalent for the test environment).

\artcomp

The script automatically detects all available runs in the suite, and runs each query for each available run.

\artout

Matching row counts emitted to stdout for verification of successful query execution, and a CSV + a Pandas dataframe logged to stdout with the query latencies.

% ----------------------------------------------
% A3 --- LLM Prompt for Agentic Workflow
% ----------------------------------------------
\newartifact

\artrel

% A3 is the LLM agent specification --- the prompt, tool wiring (ORCA controller CLI + Polars analytics), and operator-in-the-loop protocol --- used to drive the agentic debugging session against a running 4096-rank instance of A1. It supports $C_4$ by producing the localization timeline in Fig.~10.
$A_3$ is a prompt that is supplied to the LLM agent (Claude Opus 4.5 at experiment time)
to explain how to interact with an active ORCA-augmented BSP code, request telemetry, and schedule
in-situ analyses to explain anomalous behavior.
It supports $C_4$ by executing the localization workflow shown in Fig.~10.

\artexp

\begin{enumerate}
	\item (Warmup) LLM is able to interact with the running code, pause, resume, and request and analyze telemetry.
	\item (Exercise) LLM is able to progressively localize the reproduced performance anomaly.
	\item Common case telemetry is two orders of magnitude less than detailed traces.
\end{enumerate}

Caveat: continued evolution in LLM capabilities and the interactive user-guided nature of the experiment intrinsically make this demonstration challenging to reliably replicate. The anomaly demonstrated in the paper is specific to our fabric, which is relatively rare, and requires our environment for reproduction.

\arttime

\begin{itemize}
	\item Setup: 5 minutes
	\item Execution: $<$1 hour
	\item Analysis: $<$1 minute
\end{itemize}

\artin

\artinpart{Hardware}
See $A_1$.

\artinpart{Software}
See $A_1$, plus:
\begin{itemize}
	\item Claude Code with a valid Claude subscription or API key --- \url{https://www.anthropic.com/claude-code}
\end{itemize}

\artinpart{Datasets / Inputs}
A prompt \code|agentic-workflow.md| describing ORCA and documenting the experimental workflow.

\artinpart{Installation and Deployment}

Claude Code installation:
\begin{bash}
	curl -fsSL https://claude.ai/install.sh | bash
\end{bash}

Deployment:
\begin{enumerate}
	\item \code|run_orca_agentic.sh| triggers a 4096-rank AMR run with configuration parameters designed to disable anomaly mitigations and exacerbate occurrences. Simulation starts in paused mode.
	\item Claude Code must be started on another tab, and provided the IP/port of the controller interface of the ORCA controller. It is then asked to initiate the warmup exercise.
	\item After the warmup exercise is complete, the agent should be asked to start the main exercise. Minimal and guided interventions may be necessary in case it makes suboptimal steering choices.
\end{enumerate}

\artcomp

\begin{itemize}
	\item (Warmup) Agent interacts with ORCA and understands the Parquet output structure and analysis workflows.
	\item (Exercise) Agent uses high-level telemetry to detect the presence of stragglers, and progressively requests additional telemetry by writing custom in-situ flows to identify the root cause (up to \textmpiw{}, but not beyond).
\end{itemize}

\artout

\begin{itemize}
	\item (Warmup) Manipulation effects are visible in simulation stdout.
	\item (Exercise) Agent reports a complete callflow graph of Kokkos regions up to the anomalous \textmpiw{}.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  AE Appendix --- commented out, to be filled in later
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
% \appendixAE
%
% \arteval{1}
% \artin
%
% \artexpl{
% Provide instructions for installing and compiling libraries and code.
% Offer guidelines on deploying the code to resources.
% }
%
% \artcomp
%
% \artexpl{
% Describe the experiment workflow.
% If encapsulated within a workflow description or equivalent (such as a makefile or script), clearly outline the primary tasks and their interdependencies. Detail the main steps in the workflow. Merely instructing to ``Run script.sh'' is inadequate.
% }
%
% \artout
%
% \artexpl{
% \begin{itemize}
%     \item Provide a description of the expected results and a methodology for evaluating these results.
%     \item Explain how the expected results from the experiment workflow correlate with the contributions stated in the article.
%     \item For example, if the article presents results in a figure, the artifact evaluation should also produce a similar figure, depicting the same generalizable outcome. Authors must focus on these aspects to reduce the time required for others to understand and verify an artifact.
% \end{itemize}
% }
%
%
% \arteval{2}
% \artin
% \artcomp
% \artout

\end{document}
