% ! TeX root = ./main.tex

\section{Implementation}
\label{sec:orca:impl}

\newcommand{\orcaclient}{\emph{OrcaClient}}
\newcommand{\ocinit}{\texttt{Init()}}
\newcommand{\ocdest}{\texttt{Destroy()}}
\newcommand{\ocpta}{\texttt{PostTimestepAdvance()}}
\newcommand{\ocadd}{\texttt{AddRow}}

\newcommand{\topodir}{\texttt{DIRECT}}
\newcommand{\toporep}{\texttt{MPIREP}}

The core ORCA framework consists of 40K lines of code (80\% C++, 20\% Rust).
ORCA uses the Mercury~\cite{Souma2013:MercuryRPC} RPC library for on-fabric communication via
libfabric~\cite{Grun2015:libfabrica} and UCX~\cite{Shami2015:UCX} providers.
The Rust portion consists of OrcaFlow's DataFusion-based query planning and execution on each node,
while the bulk of the functionality --- bootstrap, transport, and control --- lies on the C++ side.

\subsection{Implementation Overview}

\pplushead{Integration}
Two integration surfaces exist: the \orcaclient{} shim for applications, and collectors for
profiling interfaces.

\begin{itemize}[leftmargin=*]
  \item \emph{OrcaClient:}
        The \orcaclient{} shim links with the application and exposes a \ocpta{} hook, called after every
        timestep.
        ORCA can be activated at runtime by preloading the full library, otherwise these hooks become
        no-ops.
        \ocpta{}  captures all data path work, including draining collectors, executing
        tier plans, dispatching data, and handling control messages.

  \item \emph{Collectors:}
        Streams are implemented via collectors that accumulate events locally and emit Arrow dataframes
        when drained.
        At the end of each timestep, ORCA drains all collectors and enqueues the output for analysis.
        Our implementation currently packages basic collectors for MPI~\cite{:MPI} and
        Kokkos~\cite{Trott2022:Kokkos3}.
        Additional collectors may be defined by clients.
        Collectors can also implement dynamic instrumentation via interfaces like eBPF~\cite{:eBPF} and
        DynInst~\cite{Berna2011:DynInst}.

\end{itemize}


\pplushead{Overlay}
The overlay bootstraps via a TCP-based protocol and switches to on-fabric RPCs after the initial
handshake.
While it supports additional topologies for use cases like intermediate fan-in reduction, only one
is discussed for simplicity.
% we evaluate only one for simplicity.
% only one is discussed in this paper for
% simplicity.

\pplushead{Data Path}
The data path is completely asynchronous but timestep-ordered.
Each tier executes its tier plans on dataframes in timestep order independently, and forwards
relevant data to the next tier.
Despite its asynchronous nature, flow updates across tiers also guarantee timestep-consistency.
% This also applies to flow updates across tiers
% Each overlay endpoint is assigned a 32-bit ID called \texttt{ovid} to enable uniform addressing
% across all tiers.
% Multiple topologies are supported, including \topodir{}, where all MPI ranks directly connect to
% the overlay, and \toporep{}, where ranks choose node-local representatives to reduce aggregator
% incast.
% Additional topologies can be plugged into the framework.
% For simplicity, we only discuss the \topodir{} topology in this paper.

% The ORCA overlay bootstraps via a TCP-based discovery protocol for the initial handshake with the
% controller, and then switches to on-fabric RPCs once Mercury addresses are exchanged.
% Routes are then constructed as per the selected TBON topology, and available aggregators are
% automatically assigned to ranks.

% CANDIDATE:
% \subsection{Data Path}
% \label{sec:orca:impl-data}
% % ORCA's data path is asynchronous but timestep-ordered --- each tier executes its plan on
% % dataframes in timestep order. 
% % To prevent head-of-line blocking, \tieragg{} and \tierctl{} maintain
% % a priority queue of pending GETs. 
% As discussed in \S\ref{sec:orca:des_insitu}, the data path is asynchronous but timestep-ordered --- each
% tier executes its tier plans on dataframes in timestep order.
% The head-of-line blocking concerns are mitigated via a \emph{(timestep, rank)}-ordered priority
% queue for pending GETs at \tieragg{} and \tierctl{}.
% Some key implementation aspects are described below:

% \begin{itemize}[leftmargin=*]
%   \item Each MPI rank emits one metadata RPC after every timestep
%         regardless of data volume if ORCA is enabled.
%         This is necessary for correctness, as higher tiers can only mark a timestep complete once the
%         entire subtree has reported.
%   \item Although Arrow analytics is zero-copy locally, ORCA performs a single memcpy from RDMA-registered memory to
%         allow registered memory handles to be recycled faster.
%   \item ORCA nodes are highly multithreaded and provision large thread pools for decode, compression, and
%         Parquet I/O, as sustaining O(GB/s) requires multiple cores even with efficient compression codecs such as Snappy.
%         While Arrow decode is highly efficient, costs still accumulate for complex flows at scale, with
%         O(10K) data fragments received each second.
%   \item Tier plan execution on aggregators and controller is multithreaded.
%         The same rank-partitioning strategy used to shard timestep dataframes across tiers is used to
%         partition shards across cores.
% \end{itemize}

\subsection{\twopc{}: Implementation Details}
\label{sec:orca:impl-ctl}

% The implementation consists of 
\twopc{} is implemented as separate \emph{inner} and \emph{outer} layers.
The inner layer enables a single transaction with a unique monotonic ID called \texttt{txnseq},
targeting a particular $T_{cmd}$.
% The outer layer adds features like $\Delta$-learning, automatic retries, and cross-tier
% consistency.
The outer layer adds features like automatic $\Delta$-learning (discussed in
\S\ref{sec:orca:des_ctl}) and \emph{control domains} for cross-tier consistency.
% The inner layer implements the core two-phase commit mechanism for a single transaction --- a
% transaction, with a unique ID called a \texttt{txnseq} is initiated targeting a particular
% $T_{cmd}$, and may fail if $T_{cmd} < T_{app}$.

% The outer layer adds $\Delta$-learning --- commands are automatically retried with a new $\Delta$
% until they succeed.
% \parahead{$\Delta$-learning}
% The outer \twopc{} layer automatically retries commands with an incremented $\Delta$ in case of
% aborts.
% As a result, $\Delta$ quickly converges to the natural latency of the TBON.

\parahead{Concurrency}
Correctness necessitates that only one transaction be in flight at any given time, as intermittent
failures and automatic retries can result in commands being committed out of order.
%, potentially breaking correctness.
However, a single transaction can contain an arbitrary number of commands applied as a batch.
% We find that correctness necessitates only one transaction to be inflight at any given time, as
% intermittent failures can result in result in out-of-order commits, breaking correctness.
% However, a single transaction can consist of an arbitrary number of commmands applied as a batch.

\parahead{Handling Pauses}
The control plane supports pausing and resuming the application between timesteps for basic
lifecycle management, and as primitives for more complex workflows.
An interesting case emerges from the interaction of the paused state with \twopc{} --- resume
commands in this state will be committed for a future timestep never reached by a paused
application, creating a deadlock.
Our solution recognizes that the paused state makes the application synchronous with the user and
renders \twopc{} unnecessary for consistency.
Commands in this state circumvent the protocol and are executed directly.

\parahead{Cross-tier Consistency}
Commands such as flow updates trigger tier plan updates at all tiers, and must be applied at a
consistent timestep across the entire TBON.
This is challenging as the higher tiers are asynchronous and observe application timesteps at
different times.
A tier is defined to observe a timestep when it processes its dataframes.

% Cross-tier consistency is supported via control \emph{domains} in the outer layer.
% allowing all overlay tiers to be addressable via timestep-consistent control.
% Each overlay tier corresponds to a domain, and each transaction contains a bitmask for its
% executing domains --- flow updates, for example, apply to all domains.
% Regardless of the domain, all \twopc{} commit decisions are made by \tiermpi{}, treating it as the
% single source of truth on the application timestep.
% Higher tiers snoop on protocol traffic and apply commands only if their domain bit is set.

Cross-tier consistency is supported via \emph{control domains}, which make each overlay tier
independently addressable by the control plane in a timestep-consistent manner.
Each tier corresponds to a domain, and each transaction carries a bitmask specifying which domains
it targets --- flow updates, for example, target all domains.
The inner protocol remains unchanged, with \tiermpi{} making all commit
decisions as the single source of truth on the application timestep, regardless of the domain.
Higher tiers snoop on protocol traffic and apply commands only if their domain bit is set.
% Higher tiers snoop on protocol traffic and apply commands only if their domain bit is set.
% Higher tiers snoop on protocol traffic and apply commands only if their domain bit is set.

% \begin{enumerate}[leftmargin=*]
% %   \item Each tier corresponds to a domain, and each transaction contains a bitmask corresponding to the domains
% %         to which it applies.
% %         Flow updates, for example, apply to all domains, thus all bits are set.
%   \item All transactions are first committed by \tiermpi{} regardless of the domain.
%         The bitmask is only checked at \emph{apply} time, and commands are committed only if the domain bit
%         corresponding to that tier is set.
%         As a result, \tiermpi{} acts as the single source of truth on the application timestep.
%   \item \tierctl{} and \tieragg{} act on control messages by \emph{snooping} on commit messages.
%         Commands applying to that tier are replicated internally before broadcasting the commit to the
%         subtree.
%         As the inner protocol guarantees correctness regardless of commits' arrival at \tiermpi{}
%         (\Cref{fig:orca:des-twopc}, Case C) this ensures that commands for $T_{cmd}$ are enqueued at all tiers
%         before $T_{cmd}$ is observed.
% \end{enumerate}

